---
title: "Attributes and Listening Correlations"
author: "Brian Gural"
date: "2023-09-08"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
runtime: shiny
---

# Setup
``` {r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
This document explores the relationship between various song attributes accessed from the Spotify API and my personal listening data for the top 100 artists of the past year.

## Anaylsis and data exploration
``` {r load libs, results = 'hide', include=T}
libs <- c("tidyverse", "gplots", "viridis", "factoextra", "ggiraphExtra", "shiny", "DT", "magrittr", "dplyr")
lapply(libs, require, character.only = T)
```

``` {r load data, include=T}
#lis.df <- read.csv("data/list_attr/bg_1y_t100.csv") # for non-shiny paths
lis.df <- read.csv("bg_1y_t500.csv")
```



``` {r subset to metrics, include=FALSE}
met.df <- lis.df |> 
          dplyr::select(
            track_name, min_played, artist_name, album_release_date, danceability, energy, key,
            loudness, mode, speechiness, acousticness, instrumentalness, liveness, valence, tempo)
```

### Correlation Heatmap
``` {r plot heatmap, fig.width=20, fig.height=12, echo = F}
num.df <- met.df |> 
  dplyr::select(-track_name, -artist_name, -album_release_date) 

renderPlot({cor(num.df) |> 
  gplots::heatmap.2(col = "viridis", margins = c(10,10), trace = "none") })
```

### Interactive 3D plot

#### Select variables
``` {r prep attr corrs, fig.width=12, fig.height=12, echo = F}
shiny::selectInput("x_var", "X-axis:", choices = colnames(num.df))
shiny::selectInput("y_var", "Y-axis:", choices = colnames(num.df))
shiny::selectInput("color_by", "Color points by:", choices = colnames(num.df))
```

``` {r plot attr corrs, fig.width=12, fig.height=12, echo = F}
renderPlot({
  ggplot(num.df, aes_string(x = input$x_var, y = input$y_var, color = input$color_by)) +
    geom_point(alpha = 0.3) +
    scale_color_viridis() +
    labs(title = "Correlation of music attributes") +
    theme(legend.position = "none") +
    theme_minimal()
})

```




### Principal Component Analysis (PCA)
We perform PCA to reduce the dimensionality of our data and to identify the principal components that capture the most variance in the data.

``` {r pca of attr, fig.width=12, fig.height=12, echo = F}
pca.df <- met.df |>  
  dplyr::mutate(track_artist = paste(track_name, artist_name, sep = "_")) |> 
  group_by(track_artist) |> 
  slice_head(n = 1) |> 
  column_to_rownames("track_artist") |> 
  dplyr::select(-album_release_date, -artist_name, -track_name, -mode)  %>%
  dplyr::ungroup() %>%
  dplyr::mutate(across(everything(), as.numeric))

pca.out <- pca.df |> 
           dplyr::select(-min_played) |> 
           t() |> 
           stats::prcomp() 

summary(pca.out)

pca.out <- pca.out$rotation |>
          merge(pca.df, by = 0)
```

### Interactive PCA Plot
In this section, you can choose the feature by which the points in the PCA plot will be colored.

``` {r shiny input, echo=FALSE, echo = F}
shiny::selectInput("pca_color", "Color points by:", choices = colnames(pca.df))
```

``` {r shiny plot, echo=FALSE}
renderPlot({
  ggplot(pca.out, aes_string(x = "PC1", y = "PC2", color = input$pca_color)) +
    geom_point(alpha = 0.3) +
    scale_color_viridis() +
    labs(title = "PCA of music attributes") +
    theme(legend.position = "none") +
    theme_minimal()
})
```


# Model Fitting for Clustering
We scale the data and use the silhouette method to find the optimal number of clusters for k-means clustering.

``` {r yahma_yhmn model fit, echo = F}
# Scale the data

scaled.df <- pca.df |> 
             dplyr::select(-min_played) |> 
             scale()

# Silhouette to find optimal cluster number

renderPlot({
fviz_nbclust(x = scaled.df, 
             FUNcluster = kmeans,
             method = 'silhouette')})

# its 2
```

# K-means Clustering
``` {r yahma kmeans, echo = F}
scaled.km <- kmeans(x = scaled.df, centers = 2)

# Add info back 
pca.df$cluster <- scaled.km$cluster


renderPlot({
  fviz_cluster(object = scaled.km,
             data = pca.df, labelsize = 1)
})
```



# Cluster Details
``` {r prep hist var, fig.width=12, fig.height=12, echo = F}
shiny::selectInput("hist_var", "Plot Density plot of:", choices = colnames(pca.df))
```

``` {r plot attr hist, fig.width=20, fig.height=12, echo = F}
renderPlot({
  ggplot(pca.df, aes(x = get(input$hist_var), fill = as.factor(cluster))) +
  geom_density(alpha = 0.3) +
  theme_minimal() +
  labs(x = as.character(input$hist_var), y = "Frequency", fill = "Cluster")
})
```


